Compare scANVI and SingleR annotations

### 2025-08-14 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
add final touches and render

# set up palettes
recoded_palette_df <- readr::read_tsv(recoded_palette_file)
recoded_celltype_pal <- recoded_palette_df$hex_color
names(recoded_celltype_pal) <- recoded_palette_df$harmonized_label
# for this notebook, we want just a single Unknown
recoded_celltype_pal["Unknown"] <- recoded_celltype_pal["unknown_singler"] 
recoded_celltype_pal["unknown_singler"] <- NULL
recoded_celltype_pal["unknown_consensus"] <- NULL

@@

How many cells will this remove?

@@

```{r}
scanvi_df |>
  dplyr::filter(scanvi_posterior < 0.75) |>
  dplyr::count(scanvi) |>
  dplyr::arrange(desc(n))
```

We'll therefore recode cells with posterior probabilities below 0.75 to `Unknown`.

@@

```{r}
celltype_df <- scuttle::makePerCellDF(
  merged_sce,
  use.coldata = c("sample_id", "is_xenograft", "consensus_celltype_annotation"),
  use.dimred = c("UMAP")
) |>
  # there ARE repeated barcodes so we need to keep cell_id around
  tibble::rownames_to_column("cell_id") |>
  dplyr::rename(
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2,
    consensus_annotation = consensus_celltype_annotation
  ) |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # recode consensus NAs to "Unknown"
  dplyr::mutate(
    validation_group_annotation = ifelse(
      is.na(validation_group_annotation), "Unknown", validation_group_annotation
    )
  ) |>
  dplyr::select(-consensus_annotation) |>
  # merge in the annotations
  dplyr::left_join(singler_df, by = "cell_id") |>
  dplyr::left_join(scanvi_df, by = "cell_id") |>
  # make it tidy
  tidyr::pivot_longer(
    c(validation_group_annotation, singler, scanvi), 
    names_to = "celltype_method", 
    values_to = "label"
  ) |>
  harmonize_celltypes(label, label_recoded) |>
  dplyr::mutate(label = ifelse(label == "NE", "Neuroendocrine", label)) |>
  tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE)
```

@@

```{r, fig.width = 12, fig.height = 8}
nbatlas_t_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "T cell") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

@@

nbatlas_myeloid_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "Myeloid") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

@@

nbatlas_myeloid_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "Myeloid") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

@@

```{r, fig.height = 11, fig.width = 11} 
# make a wide version for the heatmap
celltype_wide_df <- celltype_df |>
  dplyr::filter(celltype_method != "validation_group_annotation") |>
  dplyr::select(-label_recoded) |>
  tidyr::pivot_wider(
    names_from = celltype_method,
    values_from = label
  )

@@

```{r}
# ensure factors are compatible for caret
all_levels <- c(
  celltype_wide_df$scanvi, 
  celltype_wide_df$singler
) |> unique()

>>>>

# set up palettes
recoded_palette_df <- readr::read_tsv(recoded_palette_file)
recoded_celltype_pal <- recoded_palette_df$hex_color
names(recoded_celltype_pal) <- recoded_palette_df$harmonized_label
# for this notebook, we want just a single Unknown
recoded_celltype_pal["Unknown"] <- recoded_celltype_pal["unknown_singler"]

@@

What fraction of cells will this remove?

@@

```{r}
scanvi_df |>
  dplyr::filter(scanvi_posterior < 0.75) |>
  dplyr::count(scanvi) |>
  dplyr::arrange(desc(n))
```
Notably, none are Neuroendocrine (the tumor type).

We'll apply the 0.75 threshold below by recoding these cells to `Unknown`.

@@

```{r}
celltype_df <- scuttle::makePerCellDF(
  merged_sce,
  use.coldata = c("sample_id", "is_xenograft", "consensus_celltype_annotation"),
  use.dimred = c("UMAP")
) |>
  # there ARE repeated barcodes so we need to keep cell_id around
  tibble::rownames_to_column("cell_id") |>
  dplyr::rename(
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2,
    consensus_annotation = consensus_celltype_annotation
  ) |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # recode consensus NAs to "Unknown"
  dplyr::mutate(
    validation_group_annotation = ifelse(
      is.na(validation_group_annotation), "Unknown", validation_group_annotation
    )
  ) |>
  dplyr::select(-consensus_annotation) |>
  # merge in the annotations
  dplyr::left_join(singler_df, by = "cell_id") |>
  dplyr::left_join(scanvi_df, by = "cell_id") |>
  # make it tidy
  tidyr::pivot_longer(
    c(validation_group_annotation, singler, scanvi), 
    names_to = "celltype_method", 
    values_to = "label"
  ) |>
  harmonize_celltypes(label, label_recoded) |>
  dplyr::mutate(label = ifelse(label == "NE", "Neuroendocrine", label)) |>
  tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE)

# can remove now for space
rm(merged_sce)
```

@@

```{r, fig.width = 12, fig.height = 8}
# define families using the provided map
nbatlas_t_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "T cell") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

@@

@@

# we'll use these later but not for this plot
nbatlas_myeloid_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "Myeloid") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

@@

```{r, fig.height = 9, fig.width = 9} 
# make a wide version for the heatmap
celltype_wide_df <- celltype_df |>
  dplyr::filter(celltype_method != "validation_group_annotation") |>
  dplyr::select(-label_recoded) |>
  tidyr::pivot_wider(
    names_from = celltype_method,
    values_from = label
  )

@@

```{r, message=FALSE}
# ensure factors are compatible for caret
all_levels <- c(
  celltype_wide_df$scanvi, 
  celltype_wide_df$singler
) |> unique()
